Modeling Second-Language Learning from a Psychological Perspective

### 2018-03-28 doc/naaclhlt2018-latex/paper.tex
edits

Despite these promising opportunities, it remains unclear how much of current psychological
theory might be important for tasks such as the Duolingo SLAM challenge.
As one example, the field of education data mining which has attempted to build
predictive models of student learning have found excellent results using
deep neural networks~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa} 
as opposed to more traditional, and interpretable, models and approaches
that are rooted in cognitive science~\cite[e.g.,][]{Atkinson:1972rm,Atkinson:1972rz,Corbett1995bkt,Pavlik:2008rm}. Recently, Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to a more
standard ``Bayesian knowledge tracing" (BKT) models and showed
that it was possible to equate the performance of the the BKT model by additional 
features and parameters that represent core aspects of the psychology of learning
and memory such as forgetting~\cite{Khajah2016hdnt}.

Our entry to SLAM borrowed conceptually from the Khajah et al. (2016) approach. 
Specifically we used a relatively well known and powerful classification
algorithm: gradient boosting decision trees~\cite[GBDT][]{ke2017lightgbm}.  Like
deep learning, this algorithm can extract complex interactions among features,
but it has some advantages including faster training and easier integration of
diverse inputs. We then created a number of new psychologically-grounded features for the SLAM
dataset covering aspect such as user perseverance, learning processes, 
contextual factors, and cognate similarity.  After finding a model which provided the best
held-out performance on the test data set, we conducted a number of ``lesioning" studies where
we selectively removed features from the model and re-estimated the parameters in order
to assess the contribution of particular types of features.  We begin by describing our 
overall modeling approach, and then discuss some of the lessons learned from our analysis.

@@

We also encoded word length and several word characteristics gleaned from
external data sources. Past research has suggested some features of words that determine how hard words are to process, as measured by how long readers look at the word and performance in lexical decision and word identification tasks. Two such features that have somewhat independent effects are word frequency (how often the word occurs in natural language) \cite{rayner1998eye} and age-of-acquisition (the age at which children typically exhibit the word in their vocabulary) \cite{brysbaert2011effects, ferrand2011comparing}. We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161} and a feature that encoded mean age-of-acquisition (in English), derived from published age-of-acquisition norms for 30,000 English words \cite{Kuperman2012} which covered many of the words present in the dataset. Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}. As an approximate measure of linguistic similarity, we used the Levenshtein edit distance between the word tokens and their translations scaled by the length of the longer word. We found translations using Google Translate\footnote{https://cloud.google.com/translate/} and calculated the Levenshtein distance to reflect the letter-by-letter similarity of the word and its translation \cite{hyyro2001explaining}.

@@

Just as we did for word tokens, we encoded the user ID as a single-column,
high-cardinality feature. We also calculated several other user-level features that related to the ``learning type'' of a user. In particular, we encoded features that might be related to constructs such as the motivation and diligence of a user. These features could help predict how users interact with old and novel words they encounter.

%Feature 1: Motivation %compute_motivation
As a proxy for motivation, we speculated that more motivated users would complete more exercises every time they decide to use the app. To estimate this, we grouped each user's exercises into ``bursts.'' Bursts were separated by at least 24 hours \hl{todd says: did you change my code
because bursts were separate by 1 hour in my imementation}.  We used three
concrete features about these bursts, namely the mean and median number of
exercises within bursts as well as the total number of bursts of a given user (to give the model a feature related to the uncertainty in the central tendency estimates).

@@

After featurizing the training data, we trained GBDT models to minimize log loss. GBDT works by
iteratively building regression trees, each of which seeks to minimize the
residual loss from prior trees. This allows it to capture non-linear effects
and high-order interactions among features. We used the LightGBM\footnote{http://lightgbm.readthedocs.io/} implementation
of GBDT \cite{ke2017lightgbm}.

@@

The results of the lesion experiments are shown in Figure~\ref{fig:lesions}. The
models are as follows.
\begin{description}
  \item[none:] all features are included.  
  \item[user:] all user-level features, including the user ID and other
    calculated features like entropy and measures of exercise bursts, are
    removed.
  \item[userid \& user other:] only user ID or only the calculated
    features user features, respectively, are removed.
  \item[word:] Token and token root IDs; previous, next, and dependency-tree
    root word IDs; and morphological, part of speech, and dependency tree
    features are removed.
  \item[word id \& word other:] only word IDs or only other
    word features, respectively, are removed.
  \item[neighbors:] Both word IDs and other word features are removed, but
    only for the previous, next, and dependency-tree root words. Information
    about the present word is maintained.
  \item[external:] External information about the word, including corpus
    frequency, Levenshtein distance from translation, and age of acquisition,
    are removed.
  \item[temporal:] Temporal information, including number and timing of
    past encounters with the word and error tracking information, is removed.
\end{description}

@@

When approaching the Duolingo SLAM task, we hoped to leverage psychological
insights in building our model. We found that in some cases, such as using the
word's age of acquisition, this was helpful. In general, though, our model gained its
power not from hand-crafted features but from applying a powerful inference
technique (gradient boosted decision trees) to raw input about user and word IDs and
exercise features.

@@

Finally, a task in which braoder generalization was required might have
allowed psychologically inspired features to perform more competitively. In this
task, there is a large amount of labeled training data for every user and for
most words. This allow simple ID-based features to work because the past
history of a user will likely influence their future performance. However, for a
newly-encountered user or word, such an ID is useless because there will be
not relevant parameter estimate. Theory-driven features, in
contrast, can often generalize to new settings because they capture more
generic aspects of the learning task.  For example, if we were asked to generalize to
a completely new language such as German, many parts of our model would falter
but word frequency, age of acquisition, and Levenshtein distance to
first-language translation would still likely prove to be features which
have high predictive utility.

>>>>

Despite these promising opportunities, it remains unclear how much of current psychological
theory might be important for tasks such as the Duolingo SLAM challenge.
In the field of education data mining, researchers trying to build
predictive models of student learning have typically relied on traditional, and interpretable, models and approaches
that are rooted in cognitive science~\cite[e.g.,][]{Atkinson:1972rm,Atkinson:1972rz,Corbett1995bkt,Pavlik:2008rm}.
However, a recent paper found that state-of-the-art results could be achieved using deep neural networks with
little or no cognitive theory built in~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa} 
In response, a number of papers such as Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to more
standard ``Bayesian knowledge tracing" (BKT) models and showed
that it was possible to equate the performance of the the BKT model by additional 
features and parameters that represent core aspects of the psychology of learning
and memory such as forgetting and individual abilities~\cite{Khajah2016hdnt}. An ongoing debate remains in this community 
whether using flexible models with lots of data can improve over more heavily structured, theory-based models \cite{Tang:2016:DNN, xiong2016going, zhang2017incorporating}.

We positioned our entry to SLAM somewhat in between the above mentioned approaches by using 
psychologically inspired features in a fairly flexible model. 
Specifically we used a relatively well known and powerful classification
algorithm: gradient boosting decision trees~\cite[GBDT][]{ke2017lightgbm}.  Like
deep learning, this algorithm can extract complex interactions among features,
but it has some advantages including faster training and easier integration of
diverse inputs. We then created a number of new psychologically-grounded features for the SLAM
dataset covering aspect such as user perseverance, learning processes, 
contextual factors, and cognate similarity.  After finding a model which provided the best
held-out performance on the test data set, we conducted a number of ``lesioning" studies where
we selectively removed features from the model and re-estimated the parameters in order
to assess the contribution of particular types of features.  We begin by describing our 
overall modeling approach, and then discuss some of the lessons learned from our analysis.

@@

We also encoded word length and several word characteristics gleaned from
external data sources. Past research in psychology has suggested some features of words that determine how hard the word are to process, as measured by how long readers look at the word and performance in lexical decision and word identification tasks. Two such features that have somewhat independent effects are word frequency (how often the word occurs in natural language) \cite{rayner1998eye} and age-of-acquisition (the age at which children typically exhibit the word in their vocabulary) \cite{brysbaert2011effects, ferrand2011comparing}. We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161} and a feature that encoded mean age-of-acquisition (in English), derived from published age-of-acquisition norms for 30,000 English words \cite{Kuperman2012} which covered many of the words present in the dataset. Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}. As an approximate measure of linguistic similarity, we used the Levenshtein edit distance between the word tokens and their translations scaled by the length of the longer word. We found translations using Google Translate\footnote{https://cloud.google.com/translate/} and calculated the Levenshtein distance to reflect the letter-by-letter similarity of the word and its translation \cite{hyyro2001explaining}.

@@

Just as we did for word tokens, we encoded the user ID as a single-column,
high-cardinality feature. We also calculated several other user-level features that related to the ``learning type'' of a user. In particular, we encoded features that might be related to psychological constructs such as the motivation and diligence of a user. These features could help predict how users interact with old and novel words they encounter.

%Feature 1: Motivation %compute_motivation
As a proxy for motivation, we speculated that more motivated users would complete more exercises every time they decide to use the app. To estimate this, we grouped each user's exercises into ``bursts.'' Bursts were separated by at least an hour.  We used three
concrete features about these bursts, namely the mean and median number of
exercises within bursts as well as the total number of bursts of a given user (to give the model a feature related to the uncertainty in the central tendency estimates).

@@

After generating all of the features for the training data, we trained GBDT models to minimize log loss. GBDT works by
iteratively building regression trees, each of which seeks to minimize the
residual loss from prior trees. This allows it to capture non-linear effects
and high-order interactions among features. We used the LightGBM\footnote{http://lightgbm.readthedocs.io/} implementation
of GBDT \cite{ke2017lightgbm}.

@@

The results of the lesion experiments are shown in Figure~\ref{fig:lesions}. The
models are as follows.
\begin{description}
  \item[none:] all features are included.  
  \item[user:] all user-level features, including the user ID and other
    calculated features like entropy and measures of exercise bursts, are
    removed.
  \item[userid \& user other:] only user ID or only the calculated
    features user features, respectively, are removed.
  \item[word:] Token and token root IDs; previous, next, and dependency-tree
    root word IDs; and morphological, part of speech, and dependency tree
    features are removed. This does not include {\bf external} features listed below.
  \item[word id \& word other:] only word IDs or only other
    word features, respectively, are removed. {\bf word other} does not include {\bf external} features listed below.
  \item[neighbors:] Both word IDs and other word features are removed, but
    only for the previous, next, and dependency-tree root words. Information
    about the present word is maintained.
  \item[external:] External information about the word, including corpus
    frequency, Levenshtein distance from translation, and age of acquisition,
    are removed.
  \item[temporal:] Temporal information, including number and timing of
    past encounters with the word and error tracking information, is removed.
\end{description}

@@

When approaching the Duolingo SLAM task, we hoped to leverage psychological
insights in building our model. We found that in some cases, such as using the
word's age-of-acquisition, this was helpful. In general, though, our model gained its
power not from hand-crafted features but from applying a powerful inference
technique (gradient boosted decision trees) to raw input about user and word IDs and
exercise features.

@@

Finally, a task in which broader generalization was required might have
allowed psychologically inspired features to perform more competitively. In this
task, there is a large amount of labeled training data for every user and for
most words. This allow simple ID-based features to work because the past
history of a user will likely influence their future performance. However, for a
newly-encountered user or word, such an ID is useless because there will not be
enough data to estimate the relevant parameter. Theory-driven features, in
contrast, can often generalize to new settings because they capture more
generic aspects of the learning task.  For example, if we were asked to generalize to
a completely new language such as German, many parts of our model would falter
but word frequency, age of acquisition, and Levenshtein distance to
first-language translation would still likely prove to be features which
have high predictive utility.
